---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
keep_md: true
---

First loading needed libraries

```{r echo=TRUE}
library(data.table)
library(ggplot2)
library(lattice)
library(knitr)
```

## Loading and preprocessing the data

Now loading data, and converting date field to Date

```{r echo=TRUE}
if (!file.exists('./activity.csv')) {
  unzip('activity.zip', exdir='.')
}
data <- read.csv('activity.csv')
data$date <- as.Date(data$date, '%Y-%m-%d')
summary(data)
```

## What is mean total number of steps taken per day?

For now, we'll ignore NA row  
Then, preapring data for graphing steps sum by date

```{r echo=TRUE}
cleanData <- data[complete.cases(data),]
summary(cleanData)

cleanDataDT <- data.table(cleanData)
stepsByDayDT <- cleanDataDT[, sum(steps), by = date]
setnames(stepsByDayDT,"V1","stepSum")
summary(stepsByDayDT)
```

Graphing 
```{r echo=TRUE}
p <- ggplot(data=stepsByDayDT, aes(x=date, y=stepSum)) + geom_histogram(stat='identity', colour="black", fill='blue') + ggtitle('Steps Sum by Date') + xlab('Date') + ylab('Steps Sum')
print(p)
```

Next step is to see how Steps sum mean and median looks like 

```{r echo=TRUE}
mean(stepsByDayDT$stepSum)
```

```{r echo=TRUE}
median(stepsByDayDT$stepSum)
```

## What is the average daily activity pattern?

Looking at Daily mean steps activity, by measure interval 

- Preparing data
- Having a look at data
- Plotting 

```{r echo=TRUE}
dataDT <- data.table(cleanData)
head(dataDT)
intervalDT <- dataDT[, mean(steps), by = interval]
setnames(intervalDT,"V1","stepAvg")
summary(intervalDT)
plot(intervalDT$interval, intervalDT$stepAvg, type="l", main='Global average steps by interval (5 mins)', xlab= "Interval", ylab= "Average steps", col="blue" , lwd=2)
```

What is the interval having max average steps activity

```{r echo=TRUE}
#intervalDT$interval[which.max(intervalDT$stepAvg)]
sorted <- intervalDT[order(-stepAvg)]
sorted[1,]
```

## Imputing missing values

Checking missing values  

Overall check
```{r echo=TRUE}
sum(is.na(data))
```

Column by column check
```{r echo=TRUE}
colSums(is.na(data))
```

Preparing data
```{r echo=TRUE}
intervalDT <- dataDT[, mean(steps), by = interval]
setnames(intervalDT,"V1","stepAvg")
```

Setting up strategy for **Imputing** missing values  
Decision is to get overall average intervals for the step through dataset
```{r echo=TRUE}
df <- data
for (i in 1:nrow(df)) {
    if (is.na(df$steps[i])) {
        df$steps[i] <- intervalDT[which(df$interval[i] == intervalDT$interval), ]$stepAvg
    }
}
head(df)
```

Imputing missing data
```{r echo=TRUE}
dfDT <- data.table(df)
imputedData <- dfDT[, sum(steps), by = date]
setnames(imputedData,"V1","stepSum")
p <- ggplot(data=imputedData, aes(x=date, y=stepSum)) + geom_histogram(stat='identity', colour="black", fill='blue')+ ggtitle('Steps Sum by Date (mean imputed)') + xlab('Date') + ylab('Steps Sum')
print(p)
```
No real difference once data imputed, which makes sense as long as we imputed mean value

Checking for mean steps once missing data imputed
```{r echo=TRUE}
mean(imputedData$stepSum)
```
Doesn't seem to change data 

Checking for median steps once missing data imputed
```{r echo=TRUE}
median(imputedData$stepSum)
```
Doesn't seem to change data 

## Are there differences in activity patterns between weekdays and weekends?

Preparing data for Week days vs Weekend days activity comparison

- Adding new column
- Transforming this new column to Factor
```{r echo=TRUE}
Sys.setlocale("LC_TIME","en_US")
dfDT$dayType <- ifelse(weekdays(dfDT$date) %in% c('Saturday', 'Sunday'), 'weekend', 'weekday')
dfDT$dayType <- as.factor(dfDT$dayType)
summary(dfDT$dayType)
```

Graphing results
```{r echo=TRUE}
stepsByDayType <- aggregate(steps ~ interval + dayType, dfDT, mean)
p<-xyplot(stepsByDayType$steps ~ stepsByDayType$interval|stepsByDayType$dayType, main="Average Steps per Interval (Weekend and Weekday comparison)",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
print(p)
```

Obviously weekend vs week days activity is quite different

knit2html('PA1_template.Rmd')
